Viscous Erosion in Porous Media

### 2018-02-26 docs/TexNotes/Notes.tex
Still working
Working hard
NM

Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
& \tderiv{L}{t} = - \mean{\thalpha \Vn} \\
& \pderiv{\theta}{t} = \frac{1}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~(\ref{eqn:Vn2}) for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{\eps}{L^{2} \log \left(2\pi/L \right)} \ppd{\theta}{\alpha} 
+ \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}
For simplicity, we introduce the variables
\begin{align}
& M = - \mean{\thalpha \Vn} \\
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align}
Then the evolution equations can be expressed as
\begin{align}
& \tderiv{L}{t} = M \\
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{align}
We treat $\theta$ spectrally
\begin{equation}
\theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha}
\end{equation}
Then we get
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\mu = \exp \left( 4 \eps \pi^2 k^2 \int \elfun dt \right)
\end{equation}
so that the solution can be represented as
\begin{align}
\thhat_k(t) & = \frac{1}{\mu} \int^t \mu \hat{N}_k dt'
= \thhat_k(0) + \int_0^t \exp \left( -4 \eps \pi^2 k^2 \int_{t'}^t \elfun dt'' \right) \hat{N}_k \, dt'
% LONGER FORM
%& = \exp \left( -4 \eps \pi^2 k^2 \int_0^t \elfun(L) ds \right) \, 
%\int_0^t \exp \left( 4 \eps \pi^2 k^2 \int_0^s \elfun(L) ds' \right) \hat{N}_k ds \\
\end{align}
where $t'$ and $t''$ are simply dummy variables for integration. We can get the exact relationship
\begin{equation}
\thhat_k(t + \Delta t) = \thhat_k(t) 
+ \int_t^{t+\Delta t} \exp \left( -4 \eps \pi^2 k^2 \int_{t'}^{t+\Delta t} \elfun dt'' \right) \hat{N}_k \, dt'
\end{equation}

\subsection{How to scale smoothing parameters?}
We have two smoothing parameters: $\eps$ and $\sigma$. How should they scale with the simulation resolution parameters? The most important resolution parameter is $N$, the total number of points on each body, but there is also $\Delta t$ and maybe some others. Figuring out $\sigma$ is the easier part: $\sigma$ is a lengthscale and should go like $1/N$. On the other hand, $\eps$ is a diffusivity in dimensions. I did a calculation and found that $\eps$ should scale like $2/(\pi^2 N^2)$.

The calculation is based on looking at Eq.~?, where $\sigma_1 \sim 2 \pi \sqrt{\eps \Delta t}$. For the largest wavenumbers, $k_{max} \sim N/2$, we want the smoothing to beat the nonlinear term (so that overall, these modes decay). So we need $e^{-\pi^2 \eps \Delta t N^2 / 2} < 1 - C\Delta t$, which gives $\eps > 2C/(\pi^2 N^2)$, independent of $\Delta t$. If $C=1$, this gives roughly $\eps > 0.2/ N^2$. However, this analysis did not take into account how the spectrum of the nonlinear term decays (or maybe it even grows???)

@@

\subsection{Benchmark the interface evolution}
It would be good to test the interface evolution, in particular verify that the curvature-driven-flow (CDF) is 2nd order accurate in time. Initially, I thought I would be able to get exact solutions for CDF, but it turns out that the PDE (\ref{thetaEvolve}) does not reduce to simply a diffusion equation, but instead still has nonlinear terms, thus making exact solutions impractical. I also tried to recast CDF in polar coordinates for star-shapes, but still the governing PDE is not simple enough to get exact solutions. Also, in the $\theta$-$L$ formulation I tried to determine the normal velocity to make the nonlinear terms vanish, but this does not appear possible either. Instead, I can do a completely numerical convergence test.

>>>>

Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
& \tderiv{L}{t} = - \mean{\thalpha \Vn} \\
& \pderiv{\theta}{t} = \frac{1}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~(\ref{eqn:Vn2}) for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{\eps}{L^{2} \log \left(2\pi/L \right)} \ppd{\theta}{\alpha} 
+ \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}
For simplicity, we introduce the variables
\begin{align}
& M = - \mean{\thalpha \Vn} \\
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align}
Then the evolution equations can be expressed as
\begin{align}
& \tderiv{L}{t} = M \\
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{align}
We treat $\theta$ spectrally
\begin{equation}
\theta(\alpha,t) = 2 \pi \alpha + \sum_{k = -\infty}^{\infty} \thhat_k(t) e^{2\pi i k \alpha}
\end{equation}
where $2 \pi \alpha$ is the linear part and does not depend on time. Then we get
\begin{equation}
\label{thODEs}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\mu_k = \exp \left( 4 \eps \pi^2 k^2 \int \elfun dt \right)
\end{equation}
so that our system becomes
\begin{align}
\label{Leq}
& \tderiv{L}{t} = M \\
\label{theq}
& \tderiv{}{t}\left( \mu_k \theta_k \right) = \mu_k \hat{N}_k
\end{align}
We discretize Eqs.~(\ref{Leq})--(\ref{theq}) with a second-order Runge-Kutta method (the midpoint RK2). See the methods paper for details.

@@

### 2018-02-26 docs/methods/erosion_methods.tex
NM

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress is growing as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / L$, since stress is inversely proportional to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)}.
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.

@@

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align*}
  \dot{A} = \int_{\gamma} \atau \, ds \,  = \mean{\atau} L \, .
\end{align*}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$. Inserting scaling law~\eqref{eqn:stresslaw} gives
\begin{align*}
  \dot{A} \sim \frac{\mu \umax}{\log(W/L)} \, .
\end{align*}
Noting that $A \sim L^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align*}
  A (1 - \log A) = \areaconst (t_f - t)
\end{align*}
where $c_A$ is a constant.
This is an {\em implicit} relationship for how the area scales down in time during erosion. In Fig.~6a, we show the body area as a function of time measured in the simulation against the above scaling law with $\areaconst = 0.42$.

@@

To compare with this analytical prediction, we aim to measure the opening angles formed in our numerical simulations. However, our simulations lose accuracy just at the final moments before the body vanishes as a result of the shear stress blowing up (in accordance with the scaling laws discussed in Section \ref{sec:scaling}). To circumvent this problem, we could restart our simulation with a scaled-up version of the the near-final  body as the initial condition, and iterate this process until convergence. A simpler solution, though, is to remove the effect of shrinking by modifying the normal velocity $\Vn$ to have zero mean,
\begin{align*}
  \Vn \leftarrow \Vn - \mean{\Vn} \, .
\end{align*}
This modification allows erosion sculpt the body as it would normally, but artificially keeps the area fixed. We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process at equally-spaced times. Notice the body obtains a similar profile as observed previously, but with fixed area. Importantly, more numerical accuracy is now available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) shows a zoom of the body's rear stagnation point, demonstrating that the near-corner is actually smooth on a fine scale, in accordance with our numerical smoothing.

Figure~\ref{fixed_intface}(b) shows the stress distribution on the evolving body, revealing that stress becomes more uniform as time proceeds. Interestingly, the slight rise in shear near the front/rear corners that was seen in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that the effect was due to loss of accuracy.
Here, time is normalized by the same vanishing time $t_f = 1.79E-2$ as in Fig.~\ref{shrink_intface}. Curiously, about 3 multiples of $t_f$ are required to reach an approximate steady-state, which contrasts with previous studies of high-Reynold's-number erosion, where a satisfactory steady-state was usually reached before the body vanished~\cite{moo-ris-chi-zha-she2013}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/fixed_intface.pdf}
\caption{Fixed-area simulations: (a) Interfaces of a body eroding under shear stress but with the area artifically kept constant. Inset: zoom near the rear stagnation point, showing that the corner is actually smooth on a fine scale. (b) The stress distribution on the surface of the body. Over time, the stress becomes more uniform.}
\label{fixed_intface}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

The fixed-area simulations allow us to precisely measure the body's opening angle and aspect ratio as displayed in Fig.~\ref{fig:arangle}. Fig.~\ref{fig:arangle}a shows the aspect ratio, defined as the body's horizontal compared to its vertical width. The initial circular geometry has an aspect ratio of unity. As erosion causes the body to become more slender, the aspect ratio increases and tends to a value of $2.7$ in long time. Fig.~\ref{fig:arangle}b shows the body's opening angle, which we measure by fitting the tangent angle $\theta(\alpha)$ with a 7th degree polynomial over a region excluding the front/rear stagnation point (where a near-corner is approximated by a smooth curve of very high curvature). The polynomial fit is then extrapolated to the stagnation point to get an angle estimate (subject to some uncertainty, which we estimate below). As seen in Fig.~\ref{fig:arangle}a, the opening angle is initially $180^\circ$, corresponding to the initially circular geometry. Erosion causes the angle to form after a short time, and the angle becomes increasingly sharp as it converges to the value $102^\circ$ in long time. This coincides exactly with the valued predicted by our local solutions.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/arangle.pdf}
\caption{Aspect ratio and opening angle}
\label{fig:arangle}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

>>>>

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress is growing as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / L$, since stress is inversely proportional to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)} \qquad \text{as } L \to 0 \, .
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.

@@

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align*}
  \dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, .
\end{align*}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$. Inserting scaling law~\eqref{eqn:stresslaw} gives
\begin{align*}
  \dot{A} \sim \frac{\mu \umax}{\log(W/L)} \, .
\end{align*}
Noting that $A \sim L^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align*}
  A (1 - \log A) = \areaconst (t_f - t)
\end{align*}
where $c_A$ is a constant.
This is an {\em implicit} relationship for how the area scales down in time during erosion. In Fig.~6a, we show the body area as a function of time measured in the simulation against the above scaling law with $\areaconst = 0.42$.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.85 \textwidth]{./figs/fixed_intface.pdf}
\caption{Fixed-area simulations: (a) Interfaces of a body eroding under shear stress but with the area artifically kept constant. Inset: zoom near the rear stagnation point, showing that the corner is actually smooth on a fine scale. (b) The stress distribution on the surface of the body. Over time, the stress becomes more uniform.}
\label{fixed_intface}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

To compare with this analytical prediction, we aim to measure the opening angles formed in our numerical simulations. However, our simulations lose accuracy just in the final moments as the body is vanishing due to the shear stress blowing up (in accordance with the scaling laws discussed in Section \ref{sec:scaling}). To circumvent this problem, we could restart our simulation with a scaled-up version of the the near-final  body as the initial condition and iterate the process until convergence. A simpler solution, though, is to remove the effect of shrinking by modifying the normal velocity $\Vn$ to have zero mean,
\begin{align*}
  \Vn \leftarrow \Vn - \mean{\Vn} \, .
\end{align*}
This modification allows erosion sculpt the body as it normally would but artificially keeps the area fixed. We show in Fig.~\ref{fixed_intface}(a) several interfaces of this modified erosion process at equally-spaced times. Notice the body obtains a similar profile as observed previously, but with fixed area. Importantly, more numerical accuracy is now available to quantify the terminal state. The inset in Fig.~\ref{fixed_intface}(a) shows a zoom of the body's rear stagnation point, demonstrating that the near-corner is actually smooth on a fine scale, in accordance with our numerical smoothing. 
Likewise, Fig.~\ref{fixed_intface}(b) shows the stress distribution on the evolving body, revealing that stress becomes more uniform as time proceeds. Interestingly, the slight rise in shear near the front/rear corners that was seen in Fig.~\ref{shrink_intface}(b) is no longer present, suggesting that the effect was due to loss of accuracy.
Here, time is normalized by the same vanishing time $t_f = 1.79E-2$ as in Fig.~\ref{shrink_intface}. Curiously, about 3 multiples of $t_f$ are required to reach an approximate steady-state, which contrasts with previous studies of high-Reynold's-number erosion, in which a satisfactory steady-state was usually reached before the body vanished~\cite{moo-ris-chi-zha-she2013}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.85 \textwidth]{./figs/arangle.pdf}
\caption{Shape characteristics of an eroding body: (a) The aspect ratio, defined as the width to height, versus time. (b) The measured opening angle versus time. For both measurements, we show three cases of the smoothing parameters: $\eps = \sigma = 20/1024$, $\eps = \sigma = 10/1024$, and $\eps = \sigma = 5/1024$. In the most resolved simulation (the case 5/1024) the aspect ratio converges to 2.7 and the opening angle converges to the predicted value of $102^{\circ}$.}
\label{fig:arangle}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

The fixed-area simulations allow for us to precisely measure characteristics of the shape. In particular, Fig.~\ref{fig:arangle} shows measurements of the body's aspect ratio (its width to height) and the front opening angle as they vary in time. We show these measurements for three simulations in which the smoothing parameters, $\eps$ and $\sigma$, are sequentially reduced, allowing us to assess convergence in these parameters. As shown in Fig.~\ref{fig:arangle}(a), the aspect ratio begins at one for the initial circular geometry and increases in time as erosion sculpts a more slender shape. The aspect ratio reaches a long time limit of about 2.7 in the most resolved simulation. Fig.~\ref{fig:arangle}b shows the body's opening angle, which we measure by fitting the tangent angle $\theta(\alpha)$ with a 7th degree polynomial over a region excluding the front/rear stagnation point (where a near-corner is approximated by a smooth curve of very high curvature). The polynomial fit is then extrapolated to the stagnation point to obtain an angle estimate (subject to some uncertainty, which we estimate below). As seen in Fig.~\ref{fig:arangle}a, the opening angle is initially $180^\circ$, corresponding to the initially circular geometry. Erosion causes the angle to form after a short time, and the angle becomes increasingly sharp as it converges to the value $102^\circ$ in the most resolved simulation. This coincides exactly with the valued predicted by our local solutions.
